A Quantitative Genetic Linked Selection Model of the Negative Selection Process in Humans

### 2023-07-12 manuscript/manuscript.tex
theory sec changes

Throughout, we imagine a negative selection process where a continual influx of
deleterious mutations occur at a rate of $\mu$ per basepair per generation in a
conserved region of $L$ basepairs, such that the region-wide per generation
mutation rate is $U = \mu L$. Each mutation imposes a selective cost of $s$ in
heterozygotes and $2s$ in homozygotes, and fitness effects are multiplicative
across sites. When selection against a new mutation is strong, the stochastic
perturbations it experiences due to random genetic drift are weak relative to
the frequency changes caused by selection. In this case, every deleterious
mutation is destined to eventual extinction, and only the fraction of lineages
without any mutations contribute to the long-run ancestry of future
populations. The classic background selection approximates this process by
rescaling the neutral coalescent with an effective population size of $N_0 = N
\exp(-\nicefrac{U}{s})$
\parencite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1995-pt,Hudson1994-oh}.
This approximation works well for two reasons (though see
\cite{Cvijovic2018-vd,Walczak2012-fi,Nicolaisen2012-vs}). First, he number of
mutations per haplotype has has a stationary Poisson distribution with an
average equal to the deterministic mutation-selection equilibrium
\parencite{Haldane1927-ga}. Second, the time it takes for a present-day lineage
to trace their ancestry back to a mutation-free ancestor (the ``delay phase")
is negligible compared to the coalescence timescale of $N_0$ among
mutation-free ancestors (\cite{Durrett2008-ql}, p. 213; \cite{Good2014-yz}).

By contrast, when selection is weak, the frequency trajectories of both neutral
and selected sites are strongly impacted by stochastic perturbations from two
sources. First, genetic drift is due to Mendelian segregation and non-heritable
fitness differences; the magnitude of its perturbations are set by the
\emph{drift-effective} population size, $N_e$. Second, genetic draft occurs
when an allele becomes randomly associated with a fitness background that
perturbs its trajectory across generations \parencite{Neher2013-dz}. Like
genetic drift, the perturbations due to draft are directionless but overall
increase the variance in allele frequency change. Unlike drift, the impact of
the stochastic perturbations due to draft are autocorrelated and build up over
generations until the fitness background recombines off
\parencite{Robertson1961-ho,Santiago1995-hx,Buffalo2019-qs}. While draft can
generate multiple-merger coalescences that create discontinuities in allele
frequency trajectories, making diffusion approximations ill-suited
\parencite{Gillespie2000-mh,Der2011-it,Neher2013-dz}. However, when selection
is weak, draft is well-approximated as an increase in the rate of drift (or
equivalently, a reduction in effective population size). The stochastic
perturbations between selecting sites creates Hill--Robertson interference
\parencite{Hill1966-kd}.

@@

A recent model from \textcite{Santiago2016-mu} addresses these limitations by
formulating the negative selection process in terms of the equilibrium fitness
variance $V_A$ under mutation, drift, and draft.

Since their earlier work has
related the additive fitness variance in a region to the draft-effective
population experienced by a neutral allele arbitrarily far away
\parencite{Santiago1998-bs}, this can be

HERE

that derives the effective population
size experienced by a neutral allele when there is heritable fitness in the
population. This theory is in essence a model for the magnitude of stochastic
perturbations (what we think of as a \emph{draft-effective} population size)
when heritable fitness variation generates weak genetic draft. As mentioned,
perturbations due to draft are autocorrelated through time; at the individual
level Robertson considered, this is because offspring from large families tend
to beget many descendents themselves (and likewise with small families).
However, the same autocorrelation occurs at the genomic level with linkage
\parencite{Santiago1998-bs}, as the perturbations to a neutral allele's
trajectory from its particular fitness background tend to occur in the same
direction across generations until recombination.

The benefit of this quantitative genetics formulation of linked selection is
two-fold. First, as we will see, Santiago and Caballero's
(\citeyear{Santiago2016-mu}) model for the additive fitness variance under both
strong and weak negative selection is quite accurate. Second, these models of
linked selection can 
They do this through

Thus, the total additional
heritable variation in offspring number is the product of the equilibrium
additive fitness variation $V_A$ in a generation and an inflation factor
$Q_t^2$ due to the build up of autocorrelation to generation $t$. We let $Q^2
:= Q_t^2$ as $t \to \infty$. Intuitively, the product $V_A Q^2$ represents the
expected total variance in reproductive success a neutral mutation experiences
over its lifetime in a system with weak selection draft.

We can include this total additional variance into expressions for effective
population size as we would larger variance in family sizes due to
non-heritable causes \parencite{Wright1938-tv}. We define the draft-effective
population size $N_d$ as

>>>>

Throughout, we imagine a negative selection process where a continual influx of
deleterious mutations occur at a rate of $\mu$ per basepair per generation in a
conserved region of $L$ basepairs, such that the region-wide per generation
mutation rate is $U = \mu L$. Each mutation imposes a selective cost of $s$ in
heterozygotes and $2s$ in homozygotes, and fitness effects are multiplicative
across sites. When selection against a new mutation is strong, the stochastic
perturbations it experiences due to random genetic drift are weak relative to
the frequency changes caused by selection. In this case, every deleterious
mutation is destined to eventual extinction, and only the fraction of lineages
without any mutations contribute to the long-run ancestry of future
populations. The classic background selection approximates this process by
rescaling the neutral coalescent with an effective population size of $N_0 = N
\exp(-\nicefrac{U}{s})$
\parencite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1995-pt,Hudson1994-oh}.
This approximation works well for two reasons (though see
\cite{Cvijovic2018-vd,Walczak2012-fi,Nicolaisen2012-vs}). First, the number of
mutations per haplotype has has a stationary Poisson distribution with an
average equal to the deterministic mutation-selection equilibrium
\parencite{Haldane1927-ga}. Second, the time it takes for a present-day lineage
to trace their ancestry back to a mutation-free ancestor (the ``delay phase")
is negligible compared to the coalescence timescale of $N_0$ among
mutation-free ancestors (\cite{Durrett2008-ql}, p. 213; \cite{Good2014-yz}).

By contrast, when selection is weak, the frequency trajectories of both neutral
and selected sites are strongly impacted by stochastic perturbations from two
sources. First, genetic drift is due to Mendelian segregation and non-heritable
fitness differences; the magnitude of its perturbations are set by the
\emph{drift-effective} population size, $N_e$. Second, genetic draft occurs
when an allele becomes randomly associated with a fitness background that
perturbs its trajectory across generations \parencite{Neher2013-dz}. Like
genetic drift, the perturbations due to draft are directionless but overall
increase the variance in allele frequency change. Unlike drift, the impact of
the stochastic perturbations due to draft are autocorrelated and build up over
generations until the fitness background recombines off
\parencite{Robertson1961-ho,Santiago1995-hx,Buffalo2019-qs}. The stochastic
perturbations due to selection also impact selected sites, and alter their
dynamics, creating Hill--Robertson interference \parencite{Hill1966-kd}. Strong
draft can generate multiple-merger coalescences that create discontinuities in
allele frequency trajectories, making diffusion approximations ill-suited
\parencite{Gillespie2000-mh,Der2011-it,Neher2013-dz}. However, when selection
is weak, draft is well-approximated as an increase in the rate of drift (or
equivalently, a reduction in effective population size).

@@

Our work extends theory from \textcite{Santiago2016-mu}, which suggests both
strong and weak negative selection can be be accommodated with a quantitative
genetic model of linked selection. These linked selection models stem from
\textcite{Robertson1961-ho}, which in essence describes how heritable polygenic
fitness variation generates weak genetic draft. At the individual level
Robertson considered, selection generates autocorrelation as offspring from
large families tend to beget many descendents themselves (and likewise with
small families) when fitness is heritable. This same autocorrelation occurs at
the genomic level with linkage \parencite{Santiago1998-bs,Barton2000-zg}, as
the perturbations to a neutral allele's trajectory from its particular fitness
background tend to occur in the same direction across generations until the
background recombines off. These models quantify the total impact of the
autocorrelation generated by weak draft in terms of what we think of as a
\emph{draft-effective} population size $N_d$. The key insight is that in the
long-run, steady, weak draft created by the presence of additive genetic
fitness variance $V_A > 0$ acts like an extra variance in offspring number
would under pure drift \parencite{Wright1938-tv}. However, due to
autocorrelation, the cumulative effect of this fitness variance is inflated by
by a factor of $Q^2$. Intuitively, the product $V_A Q^2$ represents the
expected total variance in reproductive success a neutral mutation experiences
over its lifetime in a system with weak selection draft.

We define the draft-effective population size $N_d$ by including the total
additional variance created by heritable fitness in Wright's
(\citeyear{Wright1938-tv}) equation,
